---
title: 'Emergency Fund Modeling with NFCS Data: Evaluate Missing Data'
author: "Donnie Minnick"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r, message = FALSE}
library("dplyr")
library("knitr")
library("naniar")
library("tidyr")
library("VIM")
load("../data/nfcs_raw_data.Rdata")
```

# Quantify Missingness by Variable

```{r}
missing_summary <- nfcs_raw_data %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct)); missing_summary
```

```{r}
# View top 20 most-missing variables
head(missing_summary, 20)
```


# Visualize Missingness

```{r}
aggr(nfcs_raw_data, numbers = TRUE, sortVars = TRUE)
```

```{r}
vis_miss(nfcs_raw_data)
```

# Evaluating Missingness

There are three categories of missingness:

1. NA values, i.e. true missingness,
2. Don't know responses, which are coded as '98' in the dataset, and
3. Prefer not to say responses, which are coded as '99' in the dataset.

I can't impute values from these responses, except that in the case of Don't know (98), I did interpret it as an incorrect response when creating a composite score for financial literacy (see notebook `financial_literacy_score.Rmd` for details).

My only viable option is to treat these non-responses and NA values as missing.

# Excluding Variables Based on Missingness

The question then becomes when I exclude a variable from consideration based on missingness.

I propose the following thresholds:

* > 50% missing, exclude the variable,
* > 25% missing, evaluate more carefully and retain if important, and
* < 10-20%, acceptable threshold.

However, I also need to assess the information value of the high-missing variables.  If, for example, a variable with high missingness is very predictive, evaluate:

* Does this variable segment the population meaningfully?
* Is this variable crucial to the research question?
* Can the variable be used in a subset analysis?

# Non-Responses and Missing Values

Summary of missingness in the `nfcs_raw_data` data frame.

```{r}
response_summary <- nfcs_raw_data %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "response") %>%
  mutate(response_type = case_when(is.na(response) ~ "missing",
                                   response == 98 ~ "dk_98",
                                   response == 99 ~ "pna_99",
                                   TRUE ~ "valid")) %>%
  group_by(variable, response_type) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(percent = round(count / sum(count), 2)) %>%
  pivot_wider(names_from = response_type, 
              values_from = c(count, percent), 
              values_fill = 0) %>%
  arrange(count_valid)

kable(response_summary,
      col.names = c("Variable", 
                    "Count Missing", 
                    "Count Valid",
                    "Count 98",
                    "Count 99",
                    "Percent Missing",
                    "Percent Valid",
                    "Percent 98",
                    "Percent 99"),
      caption = "Response Summary by Question",
      format.args = list(big.mark = ","),
      align = c("l", "r", "r", "r", "r", "r", "r", "r", "r"))
```

# Greater Than 50% Missing

| ------- | --------- |
| M21_4 |
| M21_2_2015 |
| M21_1 |
| M21_3 |
| J9 |
| G35 |
| G22_2015 |
| G33 |
| E15_2015 |
| E20 |
| J6 |
| C2_2012 |
| C5_2012 |
| C3_2012 |
| C10_2012 |
| C11_2012 |

# Greater Than 20% Missing

| ------ |
| M10 |
| A14 |
| E8 |
| A10 |
| E7 |
| A21_2015 |
| M8 |
| M10 |
| M31 |
| J8 |
| F2_5 |
| F2_2 |
| F2_3 |
| M7 |
| F2_1 |
| F2_4 |
| F2_6 |






